---
title: "Global Government"
format: 
  dashboard:
    orientation: columns
    nav-buttons: [github]
    github: "https://github.com/fbbgarcia/STAT541-Project2"
    embed-resources: true
logo: images/courthouse.png
logo-alt: "Courthouse icon."
theme: cosmo
---

TO DO:
 - interactivity: add dropdown for top 5 countries in a sidebar (Kaatje)
 - clean model: get rid of estimates and plot, add description of model, distill interpretations (Jett)
 - add data scources (in footnotes?) (Franchesca)
 - increase text size in banners (Franchesca)

```{r}
#| label: dependencies

library(tidyverse)
library(gt)
library(countrycode)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
library(ggtext)
library(here)
library(ggalluvial)
library(ggfittext)
library(paletteer)
library(survival)
library(gtsummary)
library(htmltools)
library(plotly)
library(leaflet)
```

```{r}
#| label: load-data

# Load in master dataset joined from 3 sources above
# We saved this locally to avoid re-scraping on every render
democracy <- read_csv(here("Data", "democracy.csv"))

# Save global map geometries
world <- ne_countries(scale = "medium", returnclass = "sf")

# Combine democracy data with global map geometries
world_democracy <- world |>
  left_join(democracy, by = c("adm0_a3" = "country_code"))

coups <- read_tsv("http://www.uky.edu/~clthyn2/coup_data/candidate_dataset.txt")
```

# Non-Democratic Spells

## Column {width="60%"}

### Row

:::{.valuebox title="Non-Democratic Spells"}
How long does it take for countries that transition away from democracy to return to it and what factors influence this?
:::

### Row {height="50%"}

::: panel-tabset

## World Map
```{r}
#| label: map-democracy-or-not-interactive-map

# Most recent year for each country, for coloring purpose
recent <- democracy %>%
  group_by(country_code) %>%
  slice_max(year, n = 1) %>%
  ungroup() %>%
  select(country_code, year, is_democracy)

# color palette for current democracy
dem_palette <- colorFactor(
  palette = c("FALSE" = "#ee6b6e",
              "TRUE" = "steelblue"),
  domain = c(FALSE, TRUE),
  na.color = "gray"
)

# Collect centroids for markers
centroids <- st_centroid(world_democracy)
coords <- st_coordinates(centroids)

markers_coord <- centroids %>%
  mutate(
    lng = coords[,1],
    lat = coords[,2]
  )

# Regime History markers
# Keep only years where regime changes
regime_changes <- democracy %>%
  arrange(country_code, year) %>%
  group_by(country_code) %>%
  mutate(prev_regime = lag(regime_category)) %>%
  filter(year == min(year) | regime_category != prev_regime) %>%
  ungroup()

# HTML text for markers
marker_data <- regime_changes %>%
  arrange(country_code, year) %>%
  group_by(country_code, country_name) %>%
  summarise(
    regime_text = paste0(
      "<b>", first(country_name), "</b><br/>",
      paste0(year, ": ", regime_category, collapse = "<br/>")
    ),
    .groups = "drop"
  )
# join with coords for all marker data
markers_coord <- markers_coord %>%
  left_join(marker_data, by = c("adm0_a3" = "country_code"))

markers_coord <- markers_coord %>%
 filter(!is.na(regime_text))


# Leaflet interactive Plot
map <- leaflet(world_democracy) |> 
  addTiles() |> 
  #country polygons for color
  addPolygons(
    fillColor = ~dem_palette(is_democracy),
    fillOpacity = 0.7,
    color = "white",
    weight = 1
  ) |> 
  addLegend(
    position = "bottomright",
    colors = c("#ee6b6e", "steelblue"),
    values = ~is_democracy,
    title = NULL,
    labels = c("Non-Democratic", "Democratic")
  ) |> 
  addCircleMarkers(
    data = markers_coord,
    lng = ~lng,
    lat = ~lat,
    radius = 5,
    stroke = TRUE,
    color = "black",
    weight = 1,
    fillColor = "yellow",
    fillOpacity = 0.9,
    popup = ~regime_text
  )

map
```


The interactive map shows <span style='color:steelblue'><b>Democratic</b></span> and <span style='color:#ee6b6e'><b>Non-Democratic</b></span> Regimes Around the Globe. Each countries marker reveals a list of the previous regime types the country has had along with the year said regime began. Non-democratic regimes are more concentrated in regions such as Africa, the Middle East, and Southeast Asia.

## Changes

```{r}
#| label: sankey-plot

# create frequency data for sankey plot

regime_change_freq <- democracy |>
  arrange(country_code, year) |>
  group_by(country_code) |>
  mutate(post_regime = lead(regime_category)) |>
  filter((year == min(year) 
          & regime_category != post_regime) | 
           regime_category != post_regime)  |>
  mutate(regime_category = ifelse(regime_category %in% c("Parliamentary democracy",
                                                         "Mixed democratic",
                                                         "Presidential democracy",
                                                         "Civilian dictatorship",
                                                         "Military dictatorship",
                                                         "Royal dictatorship"),
                                  str_to_title(regime_category),
                                  "Other"),
         post_regime = ifelse(post_regime %in% c("Parliamentary democracy",
                                                 "Mixed democratic",
                                                 "Presidential democracy",
                                                 "Civilian dictatorship",
                                                 "Military dictatorship",
                                                 "Royal dictatorship"),
                              str_to_title(post_regime),
                              "Other")) |>
  ungroup(country_code) |>
  count(regime_category, post_regime)


# reorder regime categories from democratic to dictatorial

regime_change_freq <- regime_change_freq |>
  mutate(regime_category  = factor(regime_category,
                                   levels = c("Royal Dictatorship",
                                              "Military Dictatorship",
                                              "Civilian Dictatorship",
                                              "Presidential Democracy",
                                              "Mixed Democratic",
                                              "Parliamentary Democracy",
                                              "Other")),
         post_regime = factor(post_regime,
                              levels = c("Royal Dictatorship",
                                         "Military Dictatorship",
                                         "Civilian Dictatorship",
                                         "Presidential Democracy",
                                         "Mixed Democratic",
                                         "Parliamentary Democracy",
                                         "Other")))

# create sankey plot

cols <- c(
  "#E69F00",
  "#009E73",
  "#56B4E9",
  "#F0E442",
  "#0072B2",
  "#ee6b6e",
  "#808080"
)

regime_change_freq |>
  ggplot(aes(y = n,
             axis1 = regime_category,
             axis2 = post_regime)) +
  geom_alluvium(aes(fill = as.factor(regime_category))) +
  geom_stratum() +
  geom_fit_text(stat = "stratum",
                aes(label = str_wrap(after_stat(stratum), width = 12)),
                width = 0.25) +
  theme_minimal() +
  labs(title = "Regime Changes",
       x = "Regime",
       y = "Frequency",
       fill = "Old Regime",
       alt = "Sankey plot showing changes in political regime categories from old to new. Flows are colored by the old, or itinital, regime of the change. Regime categories for both axes include Royal Dictatorship, Military Dictatorship, Civilian Dictatorship, Presidential Democracy, Mixed Democracy, and Parliamentary Democracy (from least to most democratic). Each axis also includes an Other category comprised of non sovereign nations. Most changes to democracies are from military or civilian dictatorships.") +
  scale_x_discrete(limits = c("Old Regime", "New Regime")) +
  scale_fill_manual(values = cols)

```

Most transitions into democratic regimes originate from military or civilian dictatorship.

:::

## Column

### Row {height="50%"}

::: panel-tabset

```{r}
#| label: survival-model-km

# Survival model setup

survival_base <- democracy |>
  arrange(country_code, year) |>
  group_by(country_code) |>
  mutate(lost_democracy = is_democracy == FALSE & lag(is_democracy, default = TRUE),
         spell_id = cumsum(lost_democracy),
         next_is_democracy = lead(is_democracy))

spell_starts <- survival_base |>
  filter(lost_democracy == TRUE) |>
  select(country_code,
         spell_id,
         region,
         is_communist,
         is_multiparty)

survival <- survival_base |>
  filter(is_democracy == FALSE) |>
  group_by(country_code, spell_id) |>
  summarize(start_year = min(year),
            end_year = max(year),
            returned_to_democracy = any(next_is_democracy == TRUE, na.rm = TRUE),
            time = end_year - start_year + 1,
            .groups = "drop")

survival <- survival |>
  left_join(spell_starts, by = c("country_code", "spell_id"))

# Survival analysis model
km <- survfit(Surv(time, returned_to_democracy) ~ 1,
              data = survival)
```

## Hazard Ratio Estimates

```{r}
#| label: survival-model-cox

# Cox Proportional Hazards Model
cox_model <- coxph(Surv(time, returned_to_democracy) ~
                     region + 
                     is_communist + 
                     is_multiparty,
                   data = survival)

#summary(cox_model)

tbl_regression(cox_model, 
               exponentiate = TRUE,
               label = list(
                 is_communist ~ "Communist",
                 region ~ "Region",
                 is_multiparty ~ "Multiparty System"
                 )
               )|> 
  bold_labels() |>
  as_gt() |>
  tab_style(cell_text(weight = "bold"),
            locations = cells_column_labels(everything())) |>
  tab_caption(caption = md("Cox Proportional Hazards Model Coefficent Estimates"))
```

## Hazard Ratio Interpretations

There are 255 non-democratic spells, 178 of which return to democracy. The Cox Proportional Hazards Model significantly predicts "time to democracy" (p \< 2e-16)

- Countries in Europe & Central Asia return to democracy about **2.67 times faster** than the baseline region, East Asia & Pacific (p = 0.013).

- Countries in Latin America & the Caribbean return to democracy about **3.31 times faster** than the baseline region, East Asia & Pacific (p \< 0.001).

- Countries in the Middle East & North Africa are about **96.4% less likely** to return to democracy than the baseline region, East Asia & Pacific (p = 0.001).

- Communist regimes are about **78% less likely** to return to democracy compared to non-communist regimes (p \< 0.001).

- There is insufficient evidence to conclude that multiparty systems affect the likelihood of a non-democratic regime returning to democracy (p = 0.42).


## Plot

```{r}
plot(km,
     xlab = "Years Since Losing Democracy",
     ylab = "")

title(main = "Probability of Remaing Non-Democratic Decreases Over Time", adj = 0)
```

:::

### Row {height="50%"}

::: panel-tabset

## Regions
```{r}
#| label: governance-by-world-region-table

democracy <- democracy |>
  # Creates the region column using the countrycode package
  mutate(region = countrycode(country_code,
                              origin = 'iso3c',
                              destination = 'region',
                              # Custom matches for the codes that don't find match
                              custom_match = c('ZAR' = 'Sub-Saharan Africa',
                                               'GER' = 'Europe & Central Asia',
                                               'NUR' = 'East Asia & Pacific',
                                               'PRI' = 'Latin America & Caribbean',
                                               'ROM' = 'Europe & Central Asia'
                                               )
                              )
         )

democracy |>
  group_by(region) |>
  # Proportions of each form of governance
  summarize(n = n(),
            prop_demo = sum(is_democracy, na.rm = TRUE)/n,
            prop_pres = sum(is_presidential, na.rm = TRUE)/n,
            prop_mon = sum(is_monarchy, na.rm = TRUE)/n,
            prop_comw = sum(is_commonwealth, na.rm = TRUE)/n,
            prop_colon = sum(is_colony, na.rm = TRUE)/n,
            prop_comm = sum(is_communist, na.rm = TRUE)/n
            ) |>
  gt() |>
  cols_label(region = "World Region", 
             n = "Number of Observations",
             prop_demo = "Percent Democratic",
             prop_pres = "Percent Presidential",
             prop_mon = "Percent Monarchy",
             prop_comw = "Percent Commonwealth",
             prop_colon = "Percent Colony",
             prop_comm = "Percent Communist",
             ) |>
  tab_style(cell_text(weight = "bold"),
            locations = cells_column_labels(everything())) |>
  fmt_percent(columns = -c(n, region)) |>
  fmt_integer(columns = n) |> 
  tab_spanner(label = "Governance Type (%)",
              columns = -c(n, region)) |>
  data_color(columns = -c(n, region),
             palette = "Blues")
```

## Countries

```{r}
#| label: percent-dictatorships-per-country-table

# Calculates the number of regime changes for a given country
count_regimes <- function(country) {

  democracy |>
    filter(country_code == country) |>
    arrange(year) |>
    mutate(prev_regime = lag(regime_category)) |>
    filter(year == min(year) | regime_category != prev_regime) |>
    nrow()
  
}

# Calculates the number of regimes that were dictatorships for a given country
# Regime index of 3, 4, or 5 indicates a dictatorship
count_dictatorships <- function(country) {

  democracy |>
    filter(country_code == country) |>
    arrange(year) |>
    mutate(prev_regime = lag(regime_category)) |>
    filter(year == min(year) | regime_category != prev_regime, 
           regime_category_index >= 3) |>
    nrow()
  
}


table1 <- democracy |>
  distinct(country_code, country_name) |>
  # Uses map to iteratively apply the custom functions above and calculates the 
  # proportion of regimes that were/are dictatorships in each country
  mutate(n_regimes = map_dbl(country_code, count_regimes),
         n_dicts = map_dbl(country_code, count_dictatorships),
         prop_dict = n_dicts/n_regimes) |>
  select(country_name, n_regimes, prop_dict) |>
  arrange(desc(n_regimes)) |>
  gt() |>
  cols_label(country_name = "Country", 
             n_regimes = "Number of Regimes",
             prop_dict = "Percent Dictatorship") |>
  tab_style(cell_text(weight = "bold"),
            locations = cells_column_labels(everything())) |>
  fmt_percent(columns = prop_dict) |>
  data_color(prop_dict,
             palette = "Blues")

div(
  style = "max-height: 400px; overflow-y: auto;",
  table1
)
```

:::


# Highlight: Bolivia

## Row

:::{.valuebox title="Bolivia"}
According to Powell & Thyne (2011), Bolivia has the highest number of verified coup d'états (successful and unsuccessful) from 1950 to 2010.<br>Let's investigate the regime changes at each verified attempt.
:::

:::{.card height="80%"}

```{r}
# Identify country of interest (highest number of verified )
country_of_interest <- coups |>
  filter(verified %in% c(1, 2)) |>
  count(country, sort = TRUE) |>
  slice_head(n = 1) |>
  pull(country)

# Prepare data for plotting; join w/ coup data
plot_data <- coups |>
  filter(country == country_of_interest) |>
  drop_na(verified) |>
  arrange(year, month, day) |>
  mutate(coup_number = row_number(),
         coup_result = if_else(verified == 2, "Successful", "Unsuccessful")) |>
  left_join(democracy |>
              arrange(country_name, year) |>
              group_by(country_name) |>
              mutate(regime_after = lead(regime_category)) |>
              ungroup(),
            by = c("country" = "country_name", "year" = "year")) |>
  relocate(coup_number, coup_result, regime_category, regime_after, .after = day) |>
  mutate(hover_text = str_glue("Country: {country}<br>",
                               "Date: {month}-{day}-{year}<br>",
                               "Coup Attempt #: {coup_number}<br>",
                               "Result: {coup_result}<br>",
                               "Regime Before: {regime_category}<br>",
                               "Regime After: {regime_after}"))

# Create plot
plot <- plot_data |>
  ggplot(aes(x = year,
             y = coup_number,
             color = coup_result,
             text = hover_text,
             group = 1)) +
  geom_line(color = "gray60") +
  geom_point(size = 3) +
  scale_color_manual(values = c("Successful" = "#ee6b6e",
                                "Unsuccessful" = "steelblue")) +
  labs(title = "Bolivian Coup d'états (<span style='color:#ee6b6e;'>Successful</span> & <span style='color:steelblue;'>Unsuccessful</span>)",
       y = "Count",
       x = "Year") +
  theme_minimal() +
  theme(legend.position = "none")

# Plug into plotly
ggplotly(plot, tooltip = "text")
```

:::


